---
title: "Introduction to reducekappa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to reducekappa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# reducekappa

The `reduceKappa_wrapper` function reduces redundant pathway enrichment results by clustering similar categories based on their shared genes using kappa scores as described in the [Metascape paper](https://www.nature.com/articles/s41467-019-09234-6). It selects representative terms for each cluster and retains relevant pathway information. The function requires information from four columns to perform the full clustering process: 1) unique geneset ID, 2) descriptive name of the geneset, 3) genes returned with that category, and 4) the significance of the category. Check the default arguments and change accordingly for your data frame.

```{r}
library(reducekappa)
library(dplyr)
data(pathway_res)
```

The bundled `pathway_res` data frame contains two trials of pathway enrichment, with the `data_label` column indicating the trial. A larger example file (`pathway-res-example.txt`) is available in the package repository.

---

**Example 1: Reducing Redundant Pathways for a Single Trial**

This example processes enrichment results from a single trial and removes redundant pathways while keeping the most significant category per cluster as the representative term. To retain all results rather than only the representative, omit the `filter_representative` argument.

```{r}
pathway_reduce <- pathway_res |>
  filter(data_label == "G", FDR < 0.05) |>
  reduceKappa_wrapper(filter_representative = TRUE)

pathway_reduce |>
  select(cluster, cluster_id, cluster_term, cluster_size, Geneset.ID, Description, P.value, FDR) |>
  head(10) |>
  knitr::kable(digits = 3)
```

Check out potentially helpful attributes attached to the output:

```{r}
cat("Original terms:", attr(pathway_reduce, "n_og_terms"), "\n")
cat("Reduced terms: ", attr(pathway_reduce, "n_reduced_terms"), "\n")

attr(pathway_reduce, "cluster_info") |> head(5) |> knitr::kable()
attr(pathway_reduce, "genes_in_cluster_df") |> head(5) |> knitr::kable()
```

---

**Example 2: Clustering Across Multiple Trials**

To create consistent clusters across multiple pathway enrichment trials, use all significant pathway-associated genes as input. Set `group_slice` to the grouping column to retain the most significant term per cluster per group.

```{r}
pathway_reduce <- pathway_res |>
  filter(FDR < 0.05) |>
  reduceKappa_wrapper(
    group_slice = "data_label",
    geneset_id_col = "Geneset.ID",
    gene_col = "Genes.Returned",
    sig_col = "P.value",
    descrip_col = "Description"
  )

pathway_reduce |>
  select(cluster, cluster_id, cluster_term, cluster_size, data_label, Geneset.ID, P.value, FDR) |>
  head(10) |>
  knitr::kable(digits = 3)
```

---

**Example 3: Comparing Pathway Results Across Trials**

To compare pathway enrichment results across trials while retaining information about non-significant pathways, keep all pathways that are significant in at least one trial. Only genes from significant categories are used to define clusters.

```{r}
pathway_reduce <- pathway_res |>
  # Retain all categories that are significant in either trial
  (\(x) filter(x, Geneset.ID %in% (filter(x, FDR < 0.05) |> pull(Geneset.ID))))() |>
  # Use only genes returned from significant categories for clustering
  mutate(sig_pathway_genes = ifelse(FDR < 0.05, Genes.Returned, NA)) |>
  reduceKappa_wrapper(gene_col = "sig_pathway_genes")

pathway_reduce |>
  select(cluster, cluster_id, cluster_term, cluster_size, Geneset.ID, Description, P.value, FDR) |>
  head(10) |>
  knitr::kable(digits = 3)
```
